---
title: "Red Wine Quality Analysis"
author: "Anuja Raj"
date: "15 April 2017"
output: html_document
---

 __INTRODUCTION__ 

 I have taken Wine Quality Red dataset and apply exploratory data analysis 
 techniques to explore relationships in one variable to multiple variables and to explore 
 a selected data set for distributions, outliers, and anomalies. 
 Coding would be done on Rstudio with the help of some packages like ggplot2,knitr,dplyr,alr3,extra grid etc.


 __Univariate Plots Section__

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```
                                      
```{r echo=FALSE, results='hide',message=FALSE}
library(ggplot2)
library(GGally)
library(memisc)
library(gridExtra)
library(reshape2)
```


```{r echo=FALSE, results='hide',message=FALSE}
setwd('/home/aunja/Desktop/udacity_course_R')
red_wine <- read.csv("wineQualityReds.csv")
head(red_wine)
summary(red_wine)
```
Our dataset consists of 13 variables, with 1599 observations. 
Quality variable is discrete and the others are continuous.                                   

```{r echo=FALSE, results='hide',message=FALSE}
library(ggplot2)
ggplot(data=red_wine,aes(x=quality))+geom_bar()

```
```
   # Min.   1st Qu.   Median    Mean   3rd Qu.    Max. 
   # 3.000    5.000    6.000    5.636    6.000    8.000
```
Red wine quality is normally distributed and concentrated around 5 and 6.

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=fixed.acidity))+
  geom_histogram(bins = 30)
```

The distribution of fixed acidity is right skewed, and concentrated around 7.9


```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=volatile.acidity))+
  geom_histogram(bins = 30)
```
```
   # Min.   1st Qu.   Median    Mean    3rd Qu.    Max. 
   # 0.1200  0.3900   0.5200    0.5278  0.6400    1.5800
```
The distribution of volatile acidity seem to be unclear whether it is bimodal or unimodel,right skewed or normal. 

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=citric.acid))+
  geom_bar()
```
```
   #   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #   0.000   0.090   0.260   0.271   0.420   1.000
```
 The distribution of citric acid is not normal

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=residual.sugar))+
  geom_histogram(bins = 30)+ scale_x_log10()
grid.arrange(ggplot(red_wine, aes( x = 1, y = residual.sugar )) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'blue' ) )
            # ggplot(red_wine, aes( x   = residual.sugar  )) + 
             #      geom_histogram(bins=30 ),ncol=2)
```
```
   #   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #   0.900   1.900   2.200   2.539   2.600  15.500
```
The distribution of residual sugar is right skewed, and concentrated around 2.
There are a few outliers in the plot.

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=chlorides))+
  geom_histogram(bins = 30)+ scale_x_log10()
```
```
   #    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   # 0.01200 0.07000 0.07900 0.08747 0.09000 0.61100
```
 The distribution of chlorides is normal, and concentrated around 0.08. The plot has some outliers.

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=free.sulfur.dioxide))+
  geom_histogram(bins = 50)
```
```
   #    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #    1.00    7.00   14.00   15.87   21.00   72.00
```
 The distribution of free sulfur dioxide is right skewed and concentrated around 14

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=total.sulfur.dioxide))+
  geom_histogram(bins = 50)+scale_x_log10()
```
```
   #    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #    6.00   22.00   38.00   46.47   62.00  289.00
```
The distribution of total sulfur dioxide is right skewed and concentrated around 38.
There are a few outliers in the plot.

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=density))+
  geom_histogram(bins = 30)
```
```
   #    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #  0.9901  0.9956  0.9968  0.9967  0.9978  1.0040
```
 The distribution of density is normal and concentrated around 0.9967

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=pH))+
  geom_histogram(bins = 30)
```
```
   #   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #   2.740   3.210   3.310   3.311   3.400   4.010
```
 The distribution of pH is normal and concentrated around 3.310

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=sulphates))+
  geom_histogram(bins = 30)+scale_x_log10()
```
```
   #    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #  0.3300  0.5500  0.6200  0.6581  0.7300  2.0000
```
 The distribution of sulphates is right skewed and concentrated around 0.6581. 
 The plot has some outliers.

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data=red_wine,aes(x=alcohol))+
  geom_histogram(bins = 30)
```
```
   #    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #    8.40    9.50   10.20   10.42   11.10   14.90
```
 The distribution of alcohol is right skewed and concentrated around 10.20
 
 
 We divide the data into 3 groups: high quality group contains observations whose quality is 7 or 8, average  
 quality group contains observations whose quality is 5 or 6 and low quality group has observations whose  
 quality is 3 or 4. After examining the difference in each feature between the two groups, we see that  
 volatile acidity, density, and citric acid may have some correation 


```{r echo=FALSE, results='hide',message=FALSE}
red_wine$quality.factor <- factor(red_wine$quality)
red_wine$quality.cat <- NA
red_wine$quality.cat <- ifelse(red_wine$quality==7, 'high', NA )
red_wine$quality.cat <- ifelse(red_wine$quality==8, 'high', red_wine$quality.cat)
red_wine$quality.cat <- ifelse(red_wine$quality==3, "low", red_wine$quality.cat)
red_wine$quality.cat <- ifelse(red_wine$quality==4, "low", red_wine$quality.cat)
red_wine$quality.cat <- ifelse(red_wine$quality==5, 'average', red_wine$quality.cat)
red_wine$quality.cat <- ifelse(red_wine$quality==6, 'average', red_wine$quality.cat)
red_wine$quality.cat <- factor(red_wine$quality.cat, levels = c("low", "average", "high"))
```


 
 __Univariate Analysis__
 
 

 __Structure of the dataset__
 
 There are 1,599 red wines in the dataset with 11 features on the chemical properties of the wine.  
 fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide,  
 total.sulfur.dioxide, density, pH, sulphates, alcohol, and quality).

 __Main feature(s) of interest in the dataset__
 
 The main features in the data set are pH and quality. I’d like to determine which features are best for  
 predicting the quality of a wine. I suspect pH and some combination of the other variables can be used to  
 build a predictive model to grade the quality of wines.

 __Other features of interest in the dataset__
 
 fixed.acidity,sulphates, citric acid, and alcohol likely contribute to the quality of a wine.

 __Create new variable from existing variables in the dataset__
 
 I created a new variable called “quality.cat” which is categorically divided into “low”, “average”, and  
 “high”.                            
                          
 
 __Bivariate Plots Section__

```{r echo=FALSE, results='hide',message=FALSE}
s1 =  qplot(x = quality.cat, y = alcohol, 
      data = red_wine,
      geom = "boxplot")

s2 = qplot(x = quality.cat, y = sulphates, 
      data = red_wine,
      geom = "boxplot")

s3 = qplot(x = quality.cat, y = citric.acid, 
      data = red_wine,
      geom = "boxplot")

s4 = qplot(x = quality.cat, y = fixed.acidity, 
      data = red_wine,
      geom = "boxplot")

s5 = qplot(x = quality.cat, y = volatile.acidity, 
      data = red_wine,
      geom = "boxplot")


grid.arrange(s1, s2, s3, s4, s5, ncol = 2)

```

```{r echo=FALSE, results='hide',message=FALSE}                              
ggplot(data = red_wine,aes(x=alcohol,y=volatile.acidity))+
geom_jitter(alpha=1/10,position = position_jitter(h=0))
cor(red_wine$alcohol,red_wine$volatile.acidity)

```     

 A weak negative correlation of -0.2 exists between percent alcohol content and volatile.acidity.
                               
```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data = red_wine,aes(x=alcohol,y=residual.sugar))+
geom_point()  
cor(red_wine$alcohol,red_wine$residual.sugar)
```

 The correlation coefficient is 0.04, which indicates that there is almost no relationship between residual  
 sugar and percent alcohol content. 


```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data = red_wine,aes(x=citric.acid,y=volatile.acidity))+
geom_jitter(alpha=1/10,position = position_jitter(h=0))
cor(red_wine$citric.acid,red_wine$volatile.acidity)
```

 There is a negative correlation between citric acid and volatile acidity.


```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data = red_wine,aes(x=alcohol,y=density))+
geom_jitter(alpha=1/10,position = position_jitter(h=0))
cor(red_wine$alcohol,red_wine$density)
```

 The correlation coefficient is -0.5, so the relationship is quite clear. As percent alcohol content 
 increases, the density decreases. The reason is simple: the density of wine is lower than the density of pure  water.

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data = red_wine,aes(x=fixed.acidity,y=density))+
geom_jitter(alpha=1/10,position = position_jitter(h=0))
cor(red_wine$fixed.acidity,red_wine$density)
```

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data = red_wine,aes(x=fixed.acidity,y=citric.acid))+
 geom_jitter(alpha=1/10,position = position_jitter(h=0))
cor(red_wine$fixed.acidity,red_wine$citric.acid)
```
```{r echo=FALSE, results='hide',message=FALSE}
ggplot(data = red_wine,aes(x=fixed.acidity,y=pH))+
geom_jitter(alpha=1/10,position = position_jitter(h=0))
cor(red_wine$fixed.acidity,red_wine$pH)
```                                   

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(aes(factor(quality), 
            alcohol), 
        data = red_wine) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)
cor(red_wine$alcohol,red_wine$quality)
```  

```{r echo=FALSE, results='hide',message=FALSE}
ggplot(aes(x = alcohol, 
           y = sulphates   ), 
       data = red_wine) +
      geom_point(alpha = 0.4 ) +
      geom_smooth(method = "lm", se = FALSE,size=1)

```
                                 
                                   
 __Bivariate Analysis__
                                   
 __Talk about some of the relationships you observed in this part of the investigation. How did the feature(s)   of interest vary with other features in the dataset?__
 I observed a negative relationships between quality.cat and volatile acidity, and positive correlation  
 between quality level and alcohol. The correlation coefficient of quality.cat and citric.acid is 0.226; the  
 graph shows a weak positive relationship between quality.cat and citric acid.Alcohol and volatile acidity  
 don’t have any clear relationship between each other.

 __Did you observe any interesting relationships between the other features (not the main feature(s) of     
 interest)?__
 Yes, I observed positive relationship between density and fixed acidity, positive relationship between fixed 
 acidity and citric acid, and negative relationship between pH and fixed acidity.

 __What was the strongest relationship you found?__
 Quality is positively and strongly correlated with alcohol.


 __Multivariate Plots Section__
 
```{r echo=FALSE, results='hide',message=FALSE}                                    
ggplot(aes(x = sulphates, y = alcohol, colour = quality.cat), 
       data = red_wine) + 
  geom_point(size = 4) +
  scale_color_brewer(type = 'div', palette="Set1") +
  scale_x_continuous(lim=c(quantile(red_wine$sulphates, 0.01),
                           quantile(red_wine$sulphates, 0.99))) +
  scale_y_continuous(lim=c(quantile(red_wine$alcohol, 0.01),
                           quantile(red_wine$alcohol, 0.99))) +
  geom_point(alpha = 0.4 ) 
     
                                                 
                                     
```

 The plot reveals a clear pattern, showing most of green dots (high-quality wine) in the place where both  
 alcohol and sulphates level are high. There is also a visible range of blue dots in the middle of the plot,  
 This implies that such a combination of variables lets distinguish between different levels of medium-quality  wines (5 and 6).


```{r echo=FALSE, results='hide',message=FALSE}                                     
ggplot(aes(x = fixed.acidity, y = density, colour = quality.cat), 
       data = red_wine) + 
  geom_point(size = 4) +
  scale_color_brewer(type = 'div', palette="Set1") +
  scale_x_continuous(lim=c(quantile(red_wine$fixed.acidity, 0.01),
                           quantile(red_wine$fixed.acidity, 0.99))) +
  scale_y_continuous(lim=c(quantile(red_wine$density, 0.01),
                           quantile(red_wine$density, 0.99)))
```

 It reveals some patterns in presented data. It is visible here that the majority of green dots are  
 concentrated in the upper part, while the majority of blue dots are concentrated in the bottom part of the  
 plot. Thus, this combination of variables may be useful to distinguish medium quality wine from the high  
 quality.


```{r echo=FALSE, results='hide',message=FALSE}                                     
ggplot(aes(x = fixed.acidity, y = citric.acid, colour = quality.cat), 
       data = red_wine) + 
  geom_point(size = 4) +
  scale_color_brewer(type = 'div', palette="Set1") +
  scale_x_continuous(lim=c(quantile(red_wine$fixed.acidity, 0.01),
                           quantile(red_wine$fixed.acidity, 0.99))) +
  scale_y_continuous(lim=c(quantile(red_wine$citric.acid, 0.01),
                           quantile(red_wine$citric.acid, 0.99)))
```

 It is visible here that the  wine quality.cat, we see a positive relationship between fixed acidity and  
 citric acid.


 __Multivariate Analysis__
                                            
 From wine quality.cat with sulphates and alcohol, implies that such a combination of variables lets   
 distinguish between different levels of medium-quality wines (5 and 6).                                       From the  wine quality.cat, we see a positive relationship between fixed acidity and citric acid.  
 this combination of variables may be useful to distinguish medium quality wine from the high quality.

 __Talk about some of the relationships you observed in this part of the investigation. Were there features  
 that  strengthened each other in terms of looking at your feature(s) of interest?__
 When looking at wine quality level, we see a positive relationship between fixed acidity and citric acid


 __Final Plots and Summary__

```{r echo=FALSE, results='hide',message=FALSE}
s1 =  qplot(x = quality.cat, y = alcohol, 
      data = red_wine,
      geom = "boxplot")+xlab("quality.cat[%High,Average,Low]")+
      ylab("Alcohol[% by volume]")+
     labs(title="PLOT1")

s2 = qplot(x = quality.cat, y = citric.acid, 
      data = red_wine,
      geom = "boxplot")+xlab("quality.cat[%High,Average,Low]")+
      ylab("Citric.acid[%g/dm^3]")


s3 = qplot(x = quality.cat, y = volatile.acidity, 
      data = red_wine,
      geom = "boxplot")+xlab("quality.cat[%High,Average,Low]")+
      ylab("Volatile.acidity[%g/dm^3]")

s4 = qplot(x = quality.cat, y = pH, 
      data = red_wine,
      geom = "boxplot")+xlab("quality.cat[%High,Average,Low]")+
      ylab('pH')


grid.arrange(s1, s2, s3, s4, ncol = 2)

```

 
 
 __Description of plot1__
 
 Alcohol have high correlation with wine quality.Alcohol and citric acid are two characteristics that increase  a perceived quality of wine the most. pH and volatile acidity, on the contrary, reduce a perceived quality   
 the most.



```{r echo=FALSE, results='hide',message=FALSE}                                     
ggplot(aes(x = sulphates, y = alcohol, colour = quality.cat), 
       data = red_wine) + 
  geom_point(size = 4) +
  scale_color_brewer(type = 'div', palette="Set1") +
  scale_x_continuous(lim=c(quantile(red_wine$sulphates, 0.01),
  quantile(red_wine$sulphates, 0.99))) +
  scale_y_continuous(lim=c(quantile(red_wine$alcohol, 0.01),
  quantile(red_wine$alcohol, 0.99))) +
  geom_point(alpha = 0.4 ) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlab("Sulphates[g/dm3]")+ylab("Alcohol[% by volume]")+
  labs(title = "PLOT2")
                                     
                                     
```

 
 
 __Description of Plot2__

 It reveals some patterns in presented data. It is visible here that the majority of blue dots are  
 concentrated from middle to lower part, while the majority of green dots are concentrated in the upper part  
 of the plot. The plot indicates that the average quality of wine is more than that of low quality and high   
 quality.
 Alcohol and sulphates, together with other quality increasing characteristics, are doing the hardest job in  
 predicting red wine quality.


 __Summary__
                                       
 The wines data set contains information on 1599 wines across twelve variables from around 2009. I started by  understanding the individual variables in the data set, and then I explored interesting questions and leads  
 as I continued to make observations on plots. 
 The main features in the data set are pH and quality.
 fixed.acidity,sulphates, citric acid, and alcohol likely contribute to the quality of a wine.
 There is a positive relationship between fixed acidity and citric acid with quality.cat.
 There is a negative relationships between quality.cat and volatile acidity, and positive correlation between  quality level and alcohol. The correlation coefficient of quality.cat and citric.acid is 0.226; the graph  
 shows a weak positive relationship between quality.cat and citric acid.Alcohol and volatile acidity don’t  
 have any clear relationship between each other.
 There is a positive relationship between density and fixed acidity, positive relationship between fixed  
 acidity and citric acid, and negative relationship between pH and fixed acidity.
 There are very few wines that are rated as low or high quality. We could improve the quality of our analysis  by collecting more data, and creating more variables that may contribute to the quality of wine. Having said  that, we have successfully identified features that impact the quality of red wine, visualized their   
 relationships and summarized their statistics.
